---
title: 'Statistical report: Characteristics, outcomes and prognostic factors of hospitalized patients with heart failure with preserved and mid-range ejection fraction in comparison to patients with reduced ejection fraction – insights from the ESC Heart Failure Long-Term Registry'
author: 'Statistician: Lina Benson'
  
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    number_sections: yes
link-citations: yes
bibliography: references.bib
nocite: '@*'
---

\newpage 
\tableofcontents 
\listoftables
\listoffigures
\newpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, include = TRUE, comment = "",
  warning = FALSE, message = FALSE, fig.pos = "H",
  fig.path = "../output/figs/"
)
options(knitr.kable.NA = "")
```

```{r adjust_directory_if_needed, include=FALSE}
# Uncomment lines below if rmd file is placed in a subdirectory
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

```{r load_project}
# 1. Set options in config/global.dcf
# 2. Load packages listed in config/global.dcf
# 3. Import functions and code in lib directory

ProjectTemplate::reload.project()

# if cache should be on
cacheon <- TRUE
```             

# Data handling

## Data source

hf3_lt_fu_data_soladis_jan19.sas7bdat (from the folder DATABASE_ANALYZED.ZIP). 

## Inclusion/exclusion criteria

```{r flow}
mykable(flow, caption = "Flowchart")
mykable(outdistef, caption = "EF distribution for excluded patients", scale_down = FALSE)
```

First patient in: `r min(pdata$num_dmVisitdt)` and last patient in: `r max(pdata$num_dmVisitdt)`. 

The median age (IQR) is `r pdata %>% summarise(med = dF(median(num_age), dig = 1),
                                             q1 = dF(quantile(num_age, probs = 0.25), dig = 1),
                                             q3 = dF(quantile(num_age, probs = 0.75), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", q1, "-", q3, ")")) %>%
                                   pull(out)` and 
`r pdata %>% count(num_dmgender) %>%
  mutate(perc = dF(n / sum(n) * 100, 1)) %>%
  filter(num_dmgender == "Female") %>%
  pull(perc)`% females.  

# Statistical analysis 

## General

All analyses were performed using `r sessionInfo()$R.version$version.string` [@r]. 
The level of significance is set to 5%, two-sided. No adjustment for multiple 
comparisons were made and therefore the results should be viewed with some care.

## Missing data

For patients with missing information on the date of hospitalisation, 
`r paste0(dF(prop.table(table(pdata$outtime_hospmissing))[2] * 100, 1), "%")`, 
of all hospitalisations, the time to hospitalisation was imputed with half the time to last
follow-up. Further, times to hospitalisation that were larger than time 
to death or last follow-up were set to death or last follow-up. 

Missing data for the covariates included in the models was imputed with multiple 
imputation using mice [@mice] for 10 datasets and 10 iterations. 
Variables included in the imputation model are the same as are 
included in the cox regression. Death was included as the Nelson-Aalen estimator 
in the imputation model. 

## Descriptive statistics

Test are partly performed "overall", partly between each separate EF group 
(excluding patients from the third EF group) and partly, for categorical variables > 2 levels, 
for each separate level against all other levels grouped together as one group. 

```{r, child = "../src/base_tab.Rmd"}

```

## Outcome

All outcome analysis are performed excluding patients that died in hospital, 
were lost to follow-up or died prior to discharge (probably due to error in reporting). 

Note that patients with missing information on cause of death or hospitalisation 
are neither included in the CV death/hospitalisation or non-cv death/hospitalisation. 

The outcomes are presented with the 1 - Kaplan-Meier curves for outcomes including 
all-cause death and cumulative incidence functions for outcomes were death was considered a competing event. 
The unadjusted and adjusted HR (95% CI), p-value from Cox proportional hazards regression models are provided in the figure. 

Incidence per 1000 py was calculated with 95% Poisson confidence intervals. 
Difference between EF groups is tested with log-rank test. 

Cox proportional hazards regressions were used to evaluate possible predictors of 
outcome for all-cause mortality and first HF hospitalization. The coeffs in the 
respective EF groups were modelled using interactions with the respective variables. 
So the multivariable model is not ONE model but one model / variable since the 
interaction term needs to be included. Note that this is not the same model depicted in the graphs. 

Time is from date of discharge, or if this is missing, date of admission and 
censored at follow-up or death not defined as an event (the latter is not applicable 
to the cumulative incidence figures).  

### Assumptions

The proportional hazards assumption was investigated using the scaled Schoenfeld 
residuals (cox.zph in [@survival-package]). 
Possible outliers were visually inspected by plotting the dfbetas. 

The median (min-max) follow-up is 
`r survdata %>% summarise(med = dF(median(outtime_death / 365.25 * 12), dig = 1),
                                             min = dF(min(outtime_death / 365.25 * 12), dig = 1),
                                             max = dF(max(outtime_death / 365.25 * 12), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
                                   pull(out)` months for a total of 
                                   `r survdata %>% summarise(sumpy = dF(sum(outtime_death) / 365.25, dig = 0)) %>%
                                   pull(sumpy)` patient-years of follow-up.
                                   
```{r, child = "../src/out309012.Rmd"}

```

```{r, child = "../src/incidencerates.Rmd"}

```

\clearpage

```{r, child = "../src/ci.Rmd"}

```

\clearpage

```{r, child = "../src/cox.Rmd"}

```

\clearpage

# Reproducibility

## R session information {#sessioninfo}

`r sessionInfo()$R.version$version.string` [@r] is used, for more details see below.

```{r sessinfo}
sessionInfo()
```

## R code

The R code for all data handling and statistical analyses will be made public 
https://github.com/KIHeartFailure/characteristicsOutcomesHeartFailureESC 
after publication. It is therefore possible to link to it from the resulting article 
for increased transparency and code sharing. No data or output is stored in the 
repository. 

# References
